---
title: "resto_u"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up user-agent

```{r}
library(httr)
library(stringr)

res = GET('https://www.whatismybrowser.com/detect/what-is-my-user-agent/')
str_match_all(content(res, "text"), '<div class="value" id="detected_value">(.*?)</div>')
 
user_agent = user_agent("Mozilla/5.0 (Windows NT 10.0; Win64; x64; rv:132.0) Gecko/20100101 Firefox/132.0")
res = GET('https://www.whatismybrowser.com/detect/what-is-my-user-agent/', user_agent)
str_match_all(content(res, "text"), '<div class="value" id="detected_value">(.*?)</div>')[[1]][,2]
```
 - Select one page as an example, here we use esplanade's restaurant

```{r}
urlpage = 'https://www.crous-strasbourg.fr/restaurant/resto-u-esplanade-2/'
urlpage
```

- Get the html content using requests with your own user_agent header

- parse the html with BeautifulSoup

```{r}
res = GET(urlpage, user_agent)
soup = xml2::read_html(httr::content(res, "text"))
substr(as.character(soup), 1, 2000)
clean_soup = str_replace_all(as.character(soup),'[\\t\\r\\n\\f]','')
```

- Find out how we can get the information needed: with a right click, inspect the webpage in your navigator. See that ```'menu'``` seems to be a good option to get meals in a structured way.  

- ```'menu'``` is the html class and ```'div'```is the element: ```<div class="menu">```


```{r}
menus = str_match_all(clean_soup,  '<div class="menu">(.*?)</ul></div> *</div>')[[1]][,2]
menus
```

```{r}
meals = str_split(menus, '<div class="meal">')

# Extraire la date
date = str_match(meals[[1]][1], '<time class="menu_date_title">Menu du (.*?)</time>')[, 2]
date

dates = sapply(meals,FUN = function(x){str_match(x[1], '<time class="menu_date_title">Menu du (.*?)</time>')[, 2]})
dates

clean_meals = lapply(meals,FUN = function(x){x[-1]})
clean_meals


```

```{r}
# Initialiser un dictionnaire pour les repas
all_meals = list()
for(i in 1:length(clean_meals)) {
  meal = clean_meals[[i]]
  poles_all_infos = list()
  
  # Séparer les différentes parties des repas
  meal_foodies = str_split(meal, '</ul></li>')
  # Identifier le type de repas (Déjeuner/Dîner)
  meal_type = str_match(meal, '<div class="meal_title">(.*?)</div>')[, 2]
  
  for(j in 1:length(meal_foodies)) {
    meal_foodie = meal_foodies[[j]]
    meal_foodie = meal_foodie[-length(meal_foodie)]
    
    # Diviser entre pôle et détails
    split_parts = str_split(meal_foodie, '<ul>')
    poles = lapply(split_parts,FUN = function(split_part){str_match(split_part[1], '<li>(.+)')[, 2]})
    details = lapply(split_parts,FUN = function(split_part){str_match_all(split_part[-1], '<li>(.*?)</li>')[[1]][, 2]})
    
    
    # Exclure les pôles avec "Origine de nos viandes .."
    idx = str_detect(poles, "Origine")
    if(any(idx)){
      poles = poles[-which(idx)]
      details = details[-which(idx)]
    }
    poles_details = setNames(details,poles)
    poles_all_infos[[meal_type[[j]]]] = poles_details
  }
  
  all_meals[[dates[[i]]]] = poles_all_infos
}
all_meals
```



